Parsing Postfix log files

### 2007-11-06 doc/logparser-acronyms.tex
Replaced SMTP with \SMTP{} everywhere. Put the included tables into table environments and added a list of tables. Added the glossary and made it and the acronym list play together.
Lots of corrections, including some of Carl's. Lots more acronyms; adding those causes large diffs without appreciable changes     in the processed document.

% Warning: don't use acronyms within definitions, they don't work properly.

@@

>>>>

% % vim: set textwidth=75 spell :
% Warning: don't use acronyms within definitions, they don't work properly.

@@

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at \cite{wikipedia-sql}.
}}

\newacronym{SPF}{Sender Policy Framework}{description={
    SPF is explained fully in~\cite{openspf, wikipedia-spf}.
}}

\newacronym{ISP}{Internet Service Provider}{description={
    An ISP is a company which sells Internet access to consumers.
}}

\newacronym{LMA}{Log Mail Analyzer}{description={
    One of the other Postfix log parsers reviewed and the only example of
    published prior art~\cite{log-mail-analyser} found by the author.
}}

\newacronym{CSV}{Comma-Separated Value}{description={
    The most basic form of database available, each record is a single line
    in the file and the fields are separated by a special character,
    typically a comma or colon.
}}

### 2007-11-06 doc/logparser.tex
Lots of corrections, including some of Carl's. Lots more acronyms; adding those causes large diffs without appreciable changes     in the processed document.

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\url{http://www.postfix.org/}} Further complication arises because the set
of rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every \RBL{} returns a different
explanatory message; policy servers may log different messages depending on
the characteristics of the connection; there are many ways in which the log
lines may differ between servers, even within the same organisation when
servers are configured differently, or are running different version of
Postfix.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database.  The gathered data can then be used to optimise current
anti-spam measures, provide a baseline to test new anti-spam measures
against, or to produce statistics showing how effective those measures are.
There are numerous other uses for such data: improving server performance
by identifying troublesome destinations and reconfiguring appropriately or
dedicating transports to those destinations; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks which propagate via email; as a
base for billing customers on a shared server.  Preserving the raw data
enables users to develop a multitude of uses far beyond those conceived of
by the author.

@@

Section~\ref{database schema} partially describes the database schema used
by the parser, explaining in detail the tables used for storing the
resulting data.  The remainder of the schema is laid out in
section~\ref{rule attributes}.

Section~\ref{rules} discusses the parsing rules in detail, describing the
table in the database used to hold the rules and explaining their
structure, referring to an example rule and sample data it matches
successfully against, followed by a discussion of rule efficiency concerns.

Section~\ref{parsing-algorithm} contains the meat of the paper, describing
a naive parsing algorithm and the complications encountered which shaped
the full algorithm, followed by a comprehensive explanation of the
different stages of the algorithm, the actions taken during the execution
of the algorithm, and further complications whose solution completes the
algorithm.

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully handled by
the parsing algorithm.

Section~\ref{limitations-improvements} lists the limitations of the program
and/or the algorithm, then suggests some ways of dealing with the
limitations, with the goal of improving parsing and reproduction of the
journey a mail takes through Postfix.

Section~\ref{conclusion} describes the results of the project to date and
contains the paper's conclusion.

Appendix~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task.

@@

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix restrictions and policy servers.  The assumptions made in designing
and implementing the parser are explained, as are the conventions used in
this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
section~\refwithpage{other-parsers}), plus a review of previously published
research in this area.

@@

SQL note: $||$ is the concatenation operator in SQLite3; if the database
containing the extracted data does not support this syntax, then simply
remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
the same, just slightly less visually pleasing.

Another example is determining which restrictions are not effective: this
example shows which rules had less than 100 hits on the last log file
parsed.

@@

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to the API provided by a library: it insulates both user and
provider of the API from changes in the implementation of the API\@.  The
algorithm implemented by the parser can be improved; support can be added
for earlier or later releases of Postfix; bugs can be fixed or limitations
removed from the parser.  Statistics and/or graphs can be generated from
the database; new restrictions can be tested and the results inspected;
trends in the fight against spam can emerge from historical data saved in
the database.  Using a database simplifies writing programs which need to
interact with the data in several ways:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written as
        a shared library a new interface layer would need to be written for
        every programming language using the API\@.

@@

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of SQL, the basics are the
        same everywhere.

@@

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive and extensible optional
anti-spam restrictions, allowing an administrator to deploy those
restrictions which they judge suitable for their site's needs, rather than
a fixed set chosen by Postfix's author.  These restrictions can be
selectively applied, combined and bypassed on a per-client, per-recipient
or per-sender basis, allowing varying levels of severity and/or
permissiveness.  Postfix leverages simple lookup tables to support
arbitrarily complicated user-defined sequences of restrictions and
exceptions, with policy servers\footnote{Policy servers will be explained
shortly.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the logs generated.  While it is easy to use \texttt{grep(1)} to
determine the fate of an individual email, following the journey an email
takes through Postfix can be quite difficult.  The logs tend to follow a
90\%-10\% pattern: 90\% of the time the journey is simple, but the other
10\% of the time requires 90\% of the code.\footnote{These numbers don't
have a solid scientific basis, they're based on gut feeling from writing
and debugging the software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an MTA\@:
receive mail, send mail, local delivery of mail, etc. --- full details can
be found at~\cite{postfix-overview}.  Each log line contains the name of
the Postfix program which produced the log line, and this information is
used when determining which rules should be used to parse each log line
(see section~\refwithpage{rule characteristics} for details.)

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be handled differently from more general cases.  When the result
is another restriction Postfix will evaluate the new restriction, allowing
restrictions to be chosen based on the client IP address, HELO hostname,
sender address, recipient address, etc.  The administrator can define new
restrictions as a list of existing restrictions, allowing arbitrarily long
and complex sequences of lookups and restrictions.

@@

@@

    \item Checking Sender Policy Framework records~\cite{openspf,
        wikipedia-spf}.  SPF records specify which mail servers are allowed
        to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided \SMTP{} server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).

@@

10 other parsers have been reviewed in Appendix~\refwithpage{other-parsers}
as part of the background research for this project.  None of the reviewed
parsers perform the type of advanced parsing and log correlation described
here; all are intended to perform a specific parsing and reporting task,
rather than be a generic parser, extracting data and leaving generation of
reports from the data to other programs.  Some parsers save the data
extracted to a database but the majority discard all data once they have
finished running, making historical analysis impossible.  The other parsers
reviewed all produce a report of greater or lesser complexity and detail,
whereas the program described here doesn't attempt to produce a report at
all; that responsibility is deferred to a separate program, to be developed
later.  The parsing algorithm and program described here are designed to
enable much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from this parser: it attempts to present correlated log
data in a form suitable for a systems administrator to search using
grep(1), awk(1) and the myriad of other standard Unix text processing
utilities already available, producing a CSV file which can subsequently be
imported into a MySQL or Berkeley DB database.  The decision to support
both CSV and Berkeley DB appears to have been a limiting factor:
Comma-Separated Value is a very simple format where a record is stored in a
single line, with fields separated by a comma or other punctuation symbol.
Problems with CSV files include the difficulty in escaping separators,
providing multiple values for a field (e.g.\ multiple recipients), adding
new fields, and lack of a bundled schema describing the fields.

Berkeley DB is not an SQL database, supporting only (key, value) pairs; in
the main table the key is an integer referred to by secondary tables and
the value used is a CSV line containing all the data for that row.  The
secondary by-sender, by-recipient, by-date and by-IP tables use the
sender/recipient/date/IP as the key, and a CSV list of integers referring
to the main table is the value.  In effect this re-implements SQL foreign
keys, but without the functionality offered by even the most basic of SQL
databases (joins, ordering, searches, etc.).  It also requires custom code
to search on some combination of the above, though the authors of LMA did
provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
Berkeley DB appears to be the least useful of the three output formats

@@

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an SQL database, unlike with Berkeley
DB\@.

@@

Parsing in LMA is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\refwithpage{complications} and~\refwithpage{additional
complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
files; indeed it produces a number of warnings and no usable output.
Parsing in \parsername{} is divided in two (see
sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
detailed explanation): parsing a new log line is a trivial matter of adding
a new rule, but code changes may be required if new complications
(sections~\refwithpage{complications} and~\refwithpage{additional
complications}) must be dealt with or a new action is required
(section~\refwithpage{actions-in-detail}).

LMA does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL and
DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
have an opportunity to review the reports.  \parsername{} does not have any
bundled reports, it delegates that task to a separate, dedicated program.

In summary Log Mail Analyser appears to be a proof of concept, written to
demonstrate the point of the paper (that having this information in an
accessible fashion is useful to systems administrators), rather than a
program designed to be extensible and useful in a production environment.

@@

    \item Using a database as an application programming interface.

    \item The \SMTP{} protocol.

@@

>>>>

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\url{http://www.postfix.org/}} Further complication arises because the set
of rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every \RBL{} returns a different
explanatory message; policy servers may log different messages depending on
the characteristics of the connection; there are many ways in which the log
lines may differ between servers, even within the same organisation:
servers may be configured differently, or running different version of
Postfix.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database for later use.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying regular high volume uses (e.g.\ customer newsletters) and
restricting those uses to off-peak times; detecting virus outbreaks which
propagate via email; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

Section~\ref{database schema} describes the database schema used by the
parser, explaining in detail the tables used for storing the data gleaned
from the logs and the table that stores the rules.

Section~\ref{rules} discusses the parsing rules in detail, describing the
table in the database used to hold the rules and explaining their
structure, referring to an example rule and sample data it matches
successfully against.  This section concludes with a discussion of rule
efficiency concerns.

Section~\ref{parsing-algorithm} contains the meat of the paper, describing
a naive parsing algorithm and the complications encountered which shaped
the full algorithm, followed by a comprehensive explanation of the
different stages of the algorithm, the actions taken during execution of
the algorithm, and further complications whose solution completes the
algorithm.

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully reconstructed
by the parsing algorithm.

Section~\ref{limitations-improvements} lists the limitations of the
algorithm, then suggests some ways of dealing with the limitations, with
the goal of improving parsing and reproduction of the journey a mail takes
through Postfix.

Section~\ref{conclusion} is contains the paper's conclusion.

Appendix~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task, including why they could not be improved or
expanded upon.

@@

Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix anti-spam restrictions and policy servers.  The assumptions made in
designing and implementing the parser are explained, as are the conventions
used in this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

\SQL{} note: $||$ is the concatenation operator in SQLite3; if the database
containing the extracted data does not support this syntax, then simply
remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
the same, just slightly less visually pleasing.

Another example is determining which restrictions are not effective: this
example shows which restrictions had fewer than 100 hits on the last log
file parsed.

@@

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to any other \API{} provided by shared code: it insulates
both user and provider of the \API{} from changes in the implementation of
the \API{}\@.  The algorithm implemented by the parser can be improved;
support can be added for earlier or later releases of Postfix; bugs can be
fixed or limitations removed from the parser; these changes will not cause
the user to be negatively impacted.  Statistics and/or graphs can be
generated from the database; new restrictions can be tested and the results
inspected; trends in the fight against spam can emerge from historical data
saved in the database; the parser remains the same as the usage adapts.
Using a database simplifies writing programs which need to interact with
the data in several ways:

@@

    \item The majority of programming languages have existing code
        available allowing access to databases, so the majority of
        languages will be able to access the results obtained by running
        the parser.  If this parser was written to be utilised as shared
        code there would be little or no difficulty in using it when
        writing in the programing language it is implemented in, but every
        other programing language wishing to utilise it would require an
        interface layer to be written.  This requirement would drastically
        reduce the viability of any project wishing to build upon the work
        of this project.

@@

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
                for explanations of these terms, e.g.~\cite{wikipedia-sql}
                XXX IMPROVE.}

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of SQL, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  XXX CITE SOME RESOURCES\@.

@@

Storing the results in a database will also increase the efficiency of
using those results, as the logs need only be parsed once; indeed the
results may be used by someone with no access to the original logs.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive and extensible optional
anti-spam restrictions, allowing an administrator to deploy those
restrictions which they judge suitable for their site's needs, rather than
a fixed set chosen by Postfix's author.  These restrictions can be
selectively applied, combined and bypassed on a per-client, per-recipient
or per-sender basis, allowing varying levels of severity and/or
permissiveness.  Postfix leverages simple lookup tables to support
arbitrarily complicated user-defined sequences of restrictions and
exceptions, with policy servers\footnote{Policy servers will be explained
in section~\refwithpage{policy servers}.} as the ultimate in flexibility.
Administrators can also supply their own rejection messages to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use standard Unix text processing utilities to determine the fate of an
individual email, following the journey an email takes through Postfix can
be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
the time the journey is simple, but the other 10\% of the time requires
90\% of the code.\footnote{These numbers don't have a solid scientific
basis, they're based on gut feeling from writing and debugging the
software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an MTA\@:
receive mail, send mail, local delivery of mail, etc. --- full details can
be found at~\cite{postfix-overview}.  Each log line contains the name of
the Postfix program which produced it, and this information is used when
determining which rules should be used to parse each log line (see
section~\refwithpage{rule characteristics} for details.)

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\
\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be used as exceptions to more general cases.  When the result is
the name of another restriction Postfix will evaluate the new restriction,
allowing restrictions to be chosen based on the client IP address, HELO
hostname, sender address, recipient address, etc.  The administrator can
define new restrictions as a list of existing restrictions, allowing
arbitrarily long and complex sequences of lookups and restrictions.

@@

\label{policy servers}

@@

    \item Checking \SPF{} records~\cite{openspf,
        wikipedia-spf}.  \SPF{} records specify which mail servers are allowed
        to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).  On a well run system it is extremely unlikely that any of
        these problems will arise, though it is of course possible,
        particularly when undergoing a deluge of spam.

@@

Ten other parsers have been reviewed in
appendix~\refwithpage{other-parsers} as part of the background research for
this project.  None of the reviewed parsers perform the type of advanced
parsing and log correlation described here; all are intended to perform a
specific parsing and reporting task, rather than be a generic parser,
extracting data and leaving generation of reports from the data to other
programs.  Some parsers save the data extracted to a data store but the
majority discard all data once they have finished running, making
historical analysis impossible.  The other parsers reviewed all produce a
report of greater or lesser complexity and detail, whereas the program
described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using grep(1), awk(1)
and the myriad of other standard Unix text processing utilities already
available, producing a \CSV{} file which can subsequently be imported into
a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.

Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
value)} pairs; in the main table the key is an integer referred to by
secondary tables and the value used is a \CSV{} line containing all the
data for that row.  The secondary by-sender, by-recipient, by-date and
by-IP tables use the sender/recipient/date/IP as the key, and a \CSV{} list
of integers referring to the main table is the value.  In effect this
re-implements \SQL{} foreign keys, but without the functionality offered by
even the most basic of \SQL{} databases (joins, ordering, searches, etc.).
It also requires custom code to search on some combination of the above,
though the authors of \LMA{} did provide some queries: IP-STORY,
FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
least useful of the three output formats

@@

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an SQL database, unlike with Berkeley
DB\@; all \SQL{} databases embed the database schema into the database and
provide commands for displaying it.  Berkeley DB does not embed a schema,
as there is neither requirement nor benefit; it provides \textbf{(key,
value)} pairs, so any additional structuring of the data is imposed by the
application, thus it behoves the application to document this structure.

@@

Parsing in \LMA{} is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\refwithpage{complications} and~\refwithpage{additional
complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
files; indeed it produces a number of warnings and no usable output.
Parsing in \parsername{} is divided in two (see
sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
detailed explanation): parsing a new log line is a trivial matter of adding
a new rule, but code changes may be required if new complications
(sections~\refwithpage{complications} and~\refwithpage{additional
complications}) must be dealt with or a new action is required
(section~\refwithpage{actions-in-detail}).

\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
didn't have an opportunity to review the reports.  \parsername{} does not
have any bundled reports, it delegates that task to a separate, dedicated
program.

In summary \LMA{} appears to be a proof of concept, written to demonstrate
the point of the paper (that having this information in an accessible
fashion is useful to systems administrators), rather than a program
designed to be extensible and useful in a production environment.

@@

    \item Using a database as an \API{}.

    \item \SMTP{}.

@@

\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

\glossary{name={Phishing},description={
    Phishing~\cite{wikipedia-phishing} is an attempt to acquire information
    by masquerading as an entity trusted by the user, e.g.\ a bank.
}}

\glossary{name={Backscatter},description={
    When a spammer or worm sends mail with forged sender addresses,
    innocent sites are flooded with undeliverable mail notifications; this
    is called backscatter mail.
}}

\glossary{name={Joe job},description={
    A joe job is when spam mail is sent with a faked sender address with
    the intention of sullying the good name of the owner of the address.
    Joe jobs are a cause of backscatter, though by no means the only cause.
}}
